The algebraic hierarchy in type theory using type classes

### 2010-05-07 jar.tex
Merging changes from rough diammond.
Many changes.

\documentclass[a4paper,10pt,runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{amsmath}
\usepackage[utf8x]{inputenx}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%include polycode.fmt
%include coq2tex.tex
\arrayhs%i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\newcommand{\textPi}{\ensuremath\Pi}
\newcommand{\textSigma}{\ensuremath\Sigma}
\newcommand{\texttau}{\ensuremath\tau}
% \newcommand{\textforall}{\ensuremath\forall}
\begin{document}
\title{DRAFT: The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We\footnote{\today} propose generic design patterns to organize algebraic structures using
type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Category theory provides a further
useful abstraction. We provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
We show that the standard implementations of these are in fact instances of these type classes.
We report on the efficiency penality of these abstractions.

Finally, we use type classes to conveniently implement a quote function using the prolog-like
abilities of type class unification. We integrate this with our development of universal algebra.
\end{abstract}

\section{Introduction}
Developing formalized libraries for mathematics requires a detailed analysis of the
architecture of mathematics~\cite{C-corn,DBLP:conf/types/HaftmannW08}, similar to that of a large
software project.
The development of an algebraic hierarchy is an important aspect. Challenges include: 1. multiple
inheritance: A ring is a type with two operations. How do we inherit from the two theories of
monoids? 2. Sharing of notations. 3. Combining syntactic and semantic methods, usually this is
addressed by reflection.

Several solutions have been proposed for the Coq theorem prover~\cite{Coq,BC04}: dependent
records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.\ telescopes), packed classes~\cite{Packed} and
occasionally modules. Although the latter option is quickly discarded%
\footnote{Modules are not first class, but currently they seem to be superior for extraction to
OCaml~\cite{letouzey02}. Modules may also be seen as parametrised sections. They allow management
of namespaces, notations, hints, etc. An |Import| command imports all the constants from a
module. Thus allowing the user to type |B| instead of |A.B|. Similar functionality would be useful
for type classes.}
algebraic theories as `interfaces', as opposed to `modules', has been an important
guiding principle~\cite{Packed}.

In this fashion we implement\footnote{The sources are available
at:~\url{http://www.eelis.net/research/math-classes/}}
 algebraic theories as type classes, not instances, thus addressing the
challenges above; see Section~\ref{classes}. Type classes have recently entered
Coq~\cite{DBLP:conf/tphol/SozeauO08}. In Coq,
type classes are \emph{first class} citizens based on dependent records. The presence of dependent types makes them even more
powerful than the already succesful second class implementations in Haskell and Isabelle. Moreover,
Coq's type classes are well integrated with setoid-rewriting~\cite{Setoid-rewrite}.

We abstract from concrete algebra and formalize many-sorted universal algebra, including a
construction of free $\tau$-algebras; see Section~\ref{univ}.

% We are also concerned with efficiency.
% In fact, our aim is to use the present development as a new basis for Bishop's program in type
% theory~\cite{typesreal-article}.\marginpar{More}

So, we would like to replace data structures with more efficient ones. Concretely,
Coq provides unary, binary and machine numbers. We use type classes polymorphism to
develop a unified library for them; see Section~\ref{modul}.
The interfaces for $\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of
fractions; see Section~\ref{interfaces}. In future work we aim to base our development of the reals
on an abstract dense set, allowing us to use the efficient dyadic
rationals~\cite{boldo2009combining} as a base for exact real number computation in
Coq~\cite{Riemann,Oconnor:real}.

The paper is organised as follows: Section~\ref{classes} starts with a development of the concrete
algebraic hierarchy addressing the issues raised above. Section~\ref{univ} abstracts to universal
algebra, we develop the basic theory to the point of the first homomorphism theorem.
Section~\ref{interfaces} provides interfaces to usual datastructures by providing their universal
properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
this function is implemented in Coq's tactics language. A more convenient way is to use type class
unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparision
between type classes and canonical structures.

In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transfered
to any type theory based proof assistant supporting type classes such as
Matita~\cite{asperti2007user}.

\section{Preliminaries}
\subsection{Coq type theory}
The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types.
We will not elaborate on this, but instead refer to~\cite{Coq,BC04}.

The Coq type theory lacks quotient
types, as it would make type checking undecidable. Instead it
supports setoids (aka Bishop sets), a pair of a type and an equivalence
relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support for setoid has been added to
the system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called
|Proper|.
As we will see in Section~\ref{univ} below this functionality can also be used to work with
quotient algebras.

Coq has two kinds of sorts |Prop| for propostions and |Type| for data types. Coq
is scheduled to support proof irrelevance, every |p,q:P:Prop| are
convertible, in the future; see section~\ref{PI}.

@@

\section{To bunde or not to bundle? (Or: What to bundle?)}
We provide a rationale for our approach. We begin gently.

How should we define reflexive, transitive, and symmetric relations? In a first attempt, we
might define:
%format → = "\rightarrow"
%format Π = "\Pi"
\begin{code}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{code}
and similar for transitive and symmetric relations.

With this approach, we run into problems when defining |EquivalenceRelation| in terms of
|ReflexiveRelation| and its two siblings, because when posit three objects of type
|ReflexiveRelation|, |TransitiveRelation|, and |SymmetricRelation|, respectively, we have to
add equalities expressing that they talk about the same carrier and relation. These equalities
cause difficulties, because the three carriers and relations won't be convertible.
Technically, this leads to the research in manifest fields, i.e.\ the \emph{with} construction, to
which
we return in Section~\ref{manifest}.

As a first step we would define |ReflexiveRelation| as:
\begin{code}
  Record ReflexiveRelation (carrier: Type): Type :=
    { rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{code}

This would solve the problem for the carrier, but |EquivalenceRelation| would still need equalities
for the relation:
\begin{code}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_refl: ReflexiveRelation er_carrier
    ; er_sym: SymmetricRelation er_carrier
    ; er_trans: TransitiveRelation er_trans
    ; same_rel: rr_rel er_refl = sr_rel er_sym
    ; same_rel': rr_rel er_refl = tr_rel er_trans }.
\end{code}

A natural solution is to pull the relation out of |ReflexiveRelation| (and its siblings) as well:
\begin{code}
  Record ReflexiveRelation (carrier: Type)
                           (rel: relation carrier): Prop :=
    { rr_proof: Π x, rr_rel x x }.
\end{code}
or equivalently:
\begin{code}
  Definition ReflexiveRelation (carrier: Type)
                               (rel: relation carrier): Prop :=
    Π x, rr_rel x x.
\end{code}
This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from |Type| to |Prop| reflects the fact that |ReflexiveRelation| is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
whereas this is impossible for the latter. Since we never need to distinguish between two proofs of
a propositions, Propositions can be Opague.}
we will return to this issue in Section~\ref{PI}.
We can now define |EquivalenceRelation| without awkward equalities:
\begin{code}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: relation er_carrier
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

But looking at |EquivalenceRelation|, we see that it now essentially has the same form as the
original naive |ReflexiveRelation| record we started with. Clearly, equivalence relations
and reflexive relations are analogous concepts, and by the same argument we
conclude that |EquivalenceRelation| should really be defined as:

\begin{code}
Record EquivalenceRelation (er_carrier: Type)
                             (er_rel: relation er_carrier): Prop :=
  { er_refl: ReflexiveRelation er_carrier er_rel
  ; er_sym: SymmetricRelation er_carrier er_rel
  ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

After all, when we define a type with
two different equivalence relations on it,
such as the quotient of a setoid,
we have the same considerations as for |ReflexiveRelation| above.

At the risk of overstressing the point, let us look at how the same problem and solution keep
appearing in composite structures. Suppose we use a definition of |Monoid| that bundles the
relation. Then if we proceed to define |SemiRing| as a double |Monoid|, we face a dilemma:
\begin{code}
  Record SemiRing A: Type :=
    { mult_monoid: Monoid A 
    ; plus_monoid: Monoid A 
    ; plus_comm: Commutative (mon_op plus_monoid)
    ; mult_comm: Commutative (mon_op mult_monoid)
    ; mult_0_l: Π x: A, H
\end{code}
Do we use |mult_monoid|'s equality or |plus_monoid|'s equality in |H|?
Again, for this definition to be usable at all, we would need to add an equality expressing that the
two monoids contain the same equivalence relation. And again, the proper solution is to change
Monoid and unbundle the relation. But here, we have a choice as to how much we unbundle exactly.
(For the moment we ignore the question of whether the unit and binary operator should be unbundled
as well, focusing only on the equivalence relation and its equivalence proof.)

% We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
% \begin{code}
% Record Monoid (A: Type) (R: A → A → Prop)
%   (e: Equivalence R) : Type := ...
% \end{code}
% However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
% equivalence, we end up with two non-convertible ways to express that the relation (together with
% some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
% either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
% interchangable, and hope for automation, such as type class instance resolution, to use them
% interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
% essentially elevating them to `values', which goes against the principle of proof
% irrelevance.\marginpar{What if we add proof irrelevance?}
% BOTTOMLINE: Bundle as much as we can!
% 
% Neither of these options are very appealing. We really should just make |Monoid| take as
% parameters
% only the operations/relations. If we treat the monoid's unit and binary operation the same way,
% this
% means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
% how things ought to be.

Another benefit of our approach is that it works very smoothly when these records are in fact type
classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
structure record/class, to express that some collection of relations and
operations posess the structure as a type class constraint seems to require manifest fields; see
Section~\ref{manifest}. With
our unbundled approach, on the other hand, such type class constraints are easily expressed without
any extensions as type inhabitation queries
% (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
of types living in Prop, so that we don't even care what instances are found (thus embracing proof
irrelevance).

@@

The principles listed so far are not particularly controversial, in fact this
approach ensures maximum flexibility~\cite{Hints}. However,
there is a common perception that unbundling carriers, relations, and operations this way invariably
leads either to unmanagably big algebraic expressions (involving endless projections and/or implicit
arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
problems establishing canonical names for operations (since we no longer project them out of
structure instances all the time).

@@

Suppose we have defined |SemiGroup| in line with the aforementioned principles:
\begin{code}
Record SemiGroup (A: Type) (eq: A → A → Prop)
 (op: A → A → A): Prop := ...
\end{code}
Then our starting point is the obvious but awful:
\begin{code}
Lemma perm A eq op (P: SemiGroup A eq op) x y z:
                   eq (op (op x y) z) (op (op z y) x).
\end{code}
As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
for the semigroup operator (say, |&|) and for equality. Unfortunately, at this point the names |eq|
and |op| are just local names and there is no way to predeclare that they are to be associated with
any notation. What we really need are canonical names for these notions, so that we may fashion
notations for them.

This is a typical job for type classes, and they are used in this way in the |Clean| standard
library~\cite{Clean}. Similarly, Haskell uses type classes to overload notations.
We simply follow suit:
\begin{code}
  Class Equiv A := equiv: relation A.
  Class SemiGroupOp A := sg_op: A → A → A.

  Infix "=" := equiv: type_scope.
  Infix "&" := sg_op (at level 50, left associativity).
\end{code}
We choose to use |&| here and reserve the notation |*| for ring multiplication.

@@

We would have like to use a haskell style definition
\begin{code}
Class Equiv A := (=): relation A.
\end{code}
Unfortunately, currently this is not supported in Coq.

These single-field type classes are given special treatment in that they are not translated into
records the way classes with any other number of fields are, but are instead turned into ordinary
definitions. The effect of this is that |Equiv A| immediately reduces to |relation A|, and that the
equiv `projection' is the identity function whose argument is inferred using instance resolution.
Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with
it, notations:
\begin{code}
Record SemiGroup (A: Type) (eq: Equiv A)
                 (op: SemiGroupOp A): Prop := ...
Lemma perm A eq op (P: SemiGroup A eq op) x y z:
                 x & y & z = z & y & x.
\end{code}
We call these type classes \emph{operational type classes}. We note that operational type classes
allow us to avoid Coq's notation |scope| mechanism.

@@

On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
context. The context for |perm| looks like:
\begin{code}
  A: Type
  eq: Equiv A
  op: SemiGroupOp A
  P: SemiGroup A eq op
\end{code}

While we are not particularly worried about overly large contexts (we feel that this will most
likely not be problematic for any but the most complex formalizations), having to \emph{declare}
these entities (like |eq| and |op| for |perm|) \emph{is} a bit of a chore. Fortunately, Coq
provides a feature called implicit generalization, which is exactly what we need.
With it, we can write |perm| as follows:
\begin{code}
  Lemma perm `(SemiGroup A) x y z: x & y & z = z & y & x.
\end{code}
Thus, we have reached the mathematical ideal we aimed for.

\section{Putting it all together}\label{classes}
Combining the ideas in the previous sections we end up with a formalization of algebraic
structures where we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{code}
Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
                 {zero: RingZero A}{one: RingOne A}:Prop :=
  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
  ; semiring_plus_comm:> Commutative plus
  ; semiring_mult_comm:> Commutative mult
  ; semiring_distr:> Distribute mult plus
  ; mult_0_l: forall x, 0 * x == 0 }.
\end{code}
where e.g.\ |RingPlus| is the operation class for the semigroup operator.
The notation |:>| makes |semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
provide a ring.

\emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
can find the equivalence relation in the context.
% A similar style should be possible for, say, the |Ring| tactic, instead of
% declaring the ring structure by a separate command, we would rely on type class resolution to find
% it.
We note that |op| does not depend on the proof that |e| is an equivalence. As explained above we
use Coq's implicit quantification (|`{}|) to avoid having to write all the parameters when
\emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when
\emph{applying} a theorem. Both features are new in Coq and stem from the type class implementation.

>>>>

\documentclass[a4paper,10pt,runningheads]{llncs}
\usepackage{listings}
% The package listingsutf8 supports utf8 for \lstinputlisting
% However, we do not want to put all our code snippets in separate files.
\usepackage{url}
\usepackage{bbm}
\usepackage{amsmath}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenx}
\usepackage{comment}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
\usepackage{color}
\definecolor{dkblue}{rgb}{0,0.1,0.5}
\definecolor{lightblue}{rgb}{0,0.5,0.5}
\definecolor{dkgreen}{rgb}{0,0.4,0}
\definecolor{dk2green}{rgb}{0.4,0,0}
\definecolor{dkviolet}{rgb}{0.6,0,0.8}
\def\lstlanguagefiles{defManSSR.tex}
\lstset{language=SSR}
\begin{document}
\title{The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
We present a new formalization of the algebraic hierarchy in Coq, exploiting its new type class mechanism to make practical a solution formerly thought infeasible. Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a bare minimum. To support mathematically sound abstract interfaces for $\N$, $\Z$, and $\Q$, our formalization includes portions of category theory and multisorted universal algebra.
Algebra flourishes by the interplay between syntax and semantics. The Prolog-like
abilities of type class unification allow us to conveniently define a quote function thus facilitating the use of reflective techniques.
\end{abstract}

\section{Introduction}
The development of libraries for formalized mathematics presents many software engineering challenges~\cite{C-corn,DBLP:conf/types/HaftmannW08}, because it is far from obvious how the clean, idealized concepts from everyday mathematics should be represented using the facilities provided by concrete theorem provers and their formalisms, in a way that is both mathematically faithful and convenient to work with.

For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation.

Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a. telescopes), packed classes~\cite{Packed}, and occasionally modules. We present a new approach based entirely on Coq's new type class mechanism, and show how its features together with a key design pattern let us effectively address the challenges mentioned above.

Since we intend to use this development as a basis for constructive analysis with practical certified exact real arithmetic, an additional objective and motivation in our design is to facilitate \emph{efficient} computation. In particular, we want to be able to effortlessly swap implementations of number representations. Doing this requires that we have clean abstract interfaces, and mathematics tells us what these should look like: we represent $\N$, $\Z$, and $\Q$ as \emph{interfaces} specifying an initial semiring, an initial ring, and a field of integral fractions, respectively.

To express these elegantly and without duplication, our development\footnote{The sources are available
at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of parts of category theory and multi-sorted universal algebra, all expressed using type classes for optimum effect.

\section{The Type-Classified Algebraic Hierarchy}\label{classes}
Unlike Haskell's and Isabelle's second class type classes, Coq's type classes are first class~\cite{DBLP:conf/tphol/SozeauO08}: classes and their instances are realized as ordinary record types (``dictionaries'') and registered constants of these types.

We represent each structure in the algebraic hierarchy as a type class. This immediately leads to the familiar question of which components of the structure should become parameters of the class, and which should become fields. By far the most important design choice in our development is the decision to turn all \emph{structural} components (i.e. carriers, relations, and operations) into parameters, keeping only \emph{properties} as fields. Type classes defined in this way are essentially predicates with automatically resolved proofs.

Conventional wisdom warns that while this approach is theoretically very flexible, one risks extreme inconvenience both in having to declare and pass around all these structural components all the time, as well as in losing notations (because we no longer project named operations out of records). These legitimate concerns will be addressed in Sections~\ref{bundle}-\ref{classes}. Section~\ref{univ} abstracts to universal
algebra, we develop the basic theory to the point of the first homomorphism theorem.
Section~\ref{interfaces} provides interfaces to usual data structures by providing their universal
properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
this function is implemented in Coq's tactics language. A more convenient way is to use type class
unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparison
between type classes and canonical structures.

In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transferred
to any type theory based proof assistant supporting type classes such as
Matita~\cite{asperti2007user}.

\section{Preliminaries}
\subsection{Coq type theory}
The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types; see~\cite{Coq,BC04}.

The Coq type theory lacks quotient
types, as it would make type checking undecidable. Instead it supports setoids (aka Bishop sets), a pair of a type and an equivalence relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support for setoids has been added to the system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called \lstinline|Proper|.
As we will see in Section~\ref{univ} below this functionality can also be used to work with
quotient algebras.

Coq has two kinds of sorts \lstinline|Prop| for propositions and \lstinline|Type| for data types. Coq
is scheduled to support proof irrelevance, every \lstinline|p,q:P:Prop| are
convertible, in the future; see section~\ref{PI}.

@@

\section{To bundle or not to bundle? (Or: What to bundle?)}\label{bundle}
We provide a rationale for our approach. We begin gently.

How should we define reflexive, transitive, and symmetric relations? In a first attempt, we
might define:
% \lstinputlisting[language=SSR,inputencoding=utf8/latin1]{ReflexiveRelation.v}

\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: $\Pi$ x, rr_rel x x }.
\end{lstlisting}

and similar for transitive and symmetric relations.

With this approach, we run into problems when defining \lstinline|EquivalenceRelation| in terms of
\lstinline|ReflexiveRelation| and its two siblings, because when posit three objects of type
\lstinline|ReflexiveRelation|, \lstinline|TransitiveRelation|, and \lstinline|SymmetricRelation|, respectively, we have to
add equalities expressing that they talk about the same carrier and relation. These equalities
cause difficulties, because the three carriers and relations won't be convertible.
Technically, this leads to the research in manifest fields, i.e.\ the \emph{with} construction, to
which we return in Section~\ref{manifest}.

As a first step we would define \lstinline|ReflexiveRelation| as:
\begin{lstlisting}
  Record ReflexiveRelation (carrier: Type): Type :=
    { rr_rel: relation rr_carrier
    ; rr_proof: $\Pi$ x, rr_rel x x }.
\end{lstlisting}

This would solve the problem for the carrier, but \lstinline|EquivalenceRelation| would still need equalities for the relation:
\begin{lstlisting}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_refl: ReflexiveRelation er_carrier
    ; er_sym: SymmetricRelation er_carrier
    ; er_trans: TransitiveRelation er_trans
    ; same_rel: rr_rel er_refl = sr_rel er_sym
    ; same_rel': rr_rel er_refl = tr_rel er_trans }.
\end{lstlisting}

A natural solution is to pull the relation out of \lstinline|ReflexiveRelation| (and its siblings) as well:
\begin{lstlisting}
  Record ReflexiveRelation (carrier: Type)
                           (rel: relation carrier): Prop :=
    { rr_proof: $\Pi$ x, rr_rel x x }.
\end{lstlisting}

or equivalently:
\begin{lstlisting}
  Definition ReflexiveRelation (carrier: Type)
                               (rel: relation carrier): Prop :=
    $\Pi$ x, rr_rel x x.
\end{lstlisting}

This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from \lstinline|Type| to \lstinline|Prop| reflects the fact that \lstinline|ReflexiveRelation| is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opaque definitions. The former can be unfolded,
whereas this is impossible for the latter. Since we never need to distinguish between two proofs of
a propositions, Propositions can be Opaque.}
we will return to this issue in Section~\ref{PI}.
We can now define \lstinline|EquivalenceRelation| without awkward equalities:
\begin{lstlisting}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: relation er_carrier
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}

But looking at \lstinline|EquivalenceRelation|, we see that it now essentially has the same form as the
original naive \lstinline|ReflexiveRelation| record we started with. Clearly, equivalence relations
and reflexive relations are analogous concepts, and by the same argument we
conclude that \lstinline|EquivalenceRelation| should really be defined as:

\begin{lstlisting}
Record EquivalenceRelation (er_carrier: Type)
                             (er_rel: relation er_carrier): Prop :=
  { er_refl: ReflexiveRelation er_carrier er_rel
  ; er_sym: SymmetricRelation er_carrier er_rel
  ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}

After all, when we define a type with
two different equivalence relations on it,
such as the quotient of a setoid,
we have the same considerations as for \lstinline|ReflexiveRelation| above.

At the risk of overstressing the point, let us look at how the same problem and solution keep
appearing in composite structures. Suppose we use a definition of \lstinline|Monoid| that bundles the
relation. Then if we proceed to define \lstinline|SemiRing| as a double \lstinline|Monoid|, we face a dilemma:
\begin{lstlisting}
  Record SemiRing A: Type :=
    { mult_monoid: Monoid A 
    ; plus_monoid: Monoid A 
    ; plus_comm: Commutative (mon_op plus_monoid)
    ; mult_comm: Commutative (mon_op mult_monoid)
    ; mult_0_l: $\Pi$ x: A, H
\end{lstlisting}

Do we use \lstinline|mult_monoid|'s equality or \lstinline|plus_monoid|'s equality in \lstinline|H|?
Again, for this definition to be usable at all, we would need to add an equality expressing that the
two monoids contain the same equivalence relation. And again, the proper solution is to change
Monoid and unbundle the relation. But here, we have a choice as to how much we unbundle exactly.
(For the moment we ignore the question of whether the unit and binary operator should be unbundled
as well, focusing only on the equivalence relation and its equivalence proof.)

% We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
% \begin{lstlisting}

% Record Monoid (A: Type) (R: A → A → Prop)
%   (e: Equivalence R) : Type := ...
% \end{lstlisting}

% However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
% equivalence, we end up with two non-convertible ways to express that the relation (together with
% some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
% either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
% interchangable, and hope for automation, such as type class instance resolution, to use them
% interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
% essentially elevating them to `values', which goes against the principle of proof
% irrelevance.\marginpar{What if we add proof irrelevance?}
% BOTTOMLINE: Bundle as much as we can!
% 
% Neither of these options are very appealing. We really should just make |Monoid| take as
% parameters
% only the operations/relations. If we treat the monoid's unit and binary operation the same way,
% this
% means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
% how things ought to be.

Another benefit of our approach is that it works very smoothly when these records are in fact type
classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
structure record/class, to express that some collection of relations and
operations possess the structure as a type class constraint seems to require manifest fields; see
Section~\ref{manifest}. With
our unbundled approach, on the other hand, such type class constraints are easily expressed without
any extensions as type inhabitation queries
% (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
of types living in Prop, so that we don't even care what instances are found (thus embracing proof
irrelevance).

@@

The principles listed so far are not particularly controversial, in fact this
approach ensures maximum flexibility~\cite{Hints}. However,
there is a common perception that unbundling carriers, relations, and operations this way invariably
leads either to unmanageably big algebraic expressions (involving endless projections and/or implicit
arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
problems establishing canonical names for operations (since we no longer project them out of
structure instances all the time).

@@

Suppose we have defined \lstinline|SemiGroup| in line with the aforementioned principles:
\begin{lstlisting}
Record SemiGroup (A: Type) (eq: A $\to$ A $\to$ Prop)
 (op: A $\to$ A $\to$ A): Prop := ...
\end{lstlisting}

Then our starting point is the obvious but awful:
\begin{lstlisting}
Lemma perm A eq op (P: SemiGroup A eq op) x y z:
                   eq (op (op x y) z) (op (op z y) x).
\end{lstlisting}

As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
for the semigroup operator (say, \lstinline|&|) and for equality. Unfortunately, at this point the names \lstinline|eq| and \lstinline|op| are just local names and there is no way to predeclare that they are to be associated with any notation. What we really need are canonical names for these notions, so that we may fashion notations for them.

This is a typical job for type classes, and they are used in this way in the \lstinline|Clean| standard
library. Similarly, Haskell uses type classes to overload notations.
We simply follow suit:
\begin{lstlisting}
  Class Equiv A := equiv: relation A.
  Class SemiGroupOp A := sg_op: A $\to$ A $\to$ A.

  Infix "=" := equiv: type_scope.
  Infix "&" := sg_op (at level 50, left associativity).
\end{lstlisting}

We choose to use \lstinline|&| here and reserve the notation \lstinline|*| for ring multiplication.

@@

We would have like to use a haskell style definition
\begin{lstlisting}
Class Equiv A := (=): relation A.
\end{lstlisting}

Unfortunately, currently this is not supported in Coq.

These single-field type classes are given special treatment in that they are not translated into
records the way classes with any other number of fields are, but are instead turned into ordinary
definitions. The effect of this is that \lstinline|Equiv A| immediately reduces to \lstinline|relation A|, and that the equiv `projection' is the identity function whose argument is inferred using instance resolution. Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with it, notations:
\begin{lstlisting}
Record SemiGroup (A: Type) (eq: Equiv A)
                 (op: SemiGroupOp A): Prop := ...
Lemma perm A eq op (P: SemiGroup A eq op) x y z:
                 x & y & z = z & y & x.
\end{lstlisting}

We call these type classes \emph{operational type classes}. We note that operational type classes
allow us to avoid Coq's notation \lstinline|scope| mechanism.

@@

On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
context. The context for \lstinline|perm| looks like:
\begin{lstlisting}
  A: Type
  eq: Equiv A
  op: SemiGroupOp A
  P: SemiGroup A eq op
\end{lstlisting}

While we are not particularly worried about overly large contexts (we feel that this will most
likely not be problematic for any but the most complex formalizations), having to \emph{declare}
these entities (like \lstinline|eq| and \lstinline|op| for \lstinline|perm|) \emph{is} a bit of a chore. Fortunately, Coq provides a feature called implicit generalization, which is exactly what we need.
With it, we can write \lstinline|perm| as follows:
\begin{lstlisting}
  Lemma perm `(SemiGroup A) x y z: x & y & z = z & y & x.
\end{lstlisting}

Thus, we have reached the mathematical ideal we aimed for.

\section{Putting it all together}\label{classes}
Combining the ideas in the previous sections we end up with a formalization of algebraic
structures where we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{lstlisting}
Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
                 {zero: RingZero A}{one: RingOne A}:Prop :=
  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
  ; semiring_plus_comm:> Commutative plus
  ; semiring_mult_comm:> Commutative mult
  ; semiring_distr:> Distribute mult plus
  ; mult_0_l: forall x, 0 * x == 0 }.
\end{lstlisting}

where e.g.\ \lstinline|RingPlus| is the operation class for the semigroup operator.
The notation \lstinline|:>| makes \lstinline|semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
provide a ring.

Having argued that the \emph{all-structure-as-parameters} approach \emph{can} be made practical, we enumerate some of the benefits that make it worthwhile.

First, multiple inheritance becomes trivial: |SemiRing| inherits two |Monoid| structures on the same carrier and setoid relation, using ordinary named arguments to achieve ``manifest fields''; see section~\ref{manifest}.

Second, because our terms are small and independent and never refer to proofs, we are invulnerable to concerns about efficiency and ambiguity of projection paths that plague existing solutions, obviating the need for extensions like the proposed coercion pullbacks~\cite{Hints}.

Third, since our structural type classes are mere predicates, overlap between their instances is a non-issue. Together with the previous point, this gives us tremendous freedom to posit multiple structures on the same operations and relations, including ones derived implicitly via subclasses: by simply declaring a |SemiRing| class instance showing that a ring is a semiring, results about semirings immediately apply implicitly to any known ring, without us having to explicitly encode this relation in the hierarchy definition itself, and without needing any projection or translation of carriers or operations.

Fourth, \emph{Structure-as-parameters} helps setoid-rewriting: type class resolution
can find the equivalence relation in the context.\footnote{
A similar style should be possible for, say, the |Ring| tactic, instead of
declaring the ring structure by a separate command, we would rely on type class resolution to find it.}
We note that \lstinline|op| does not depend on the proof that \lstinline|e| is an equivalence. As explained above we use Coq's implicit quantification (\lstinline|`{}|) to avoid having to write all the parameters when \emph{stating} a theorem and Coq's maximally inserted implicit arguments to find the parameters when \emph{applying} a theorem. Both features are new in Coq and stem from the type class implementation.
